Estimation of the perceived time of presence of sources in urban acoustic environment using deep learning techniques

### 2019-06-11 paper/paperV1.tex
final ml proof read

\documentclass[11pt,a4paper]{article}
\usepackage[english]{babel}
\usepackage[T1]{fontenc}

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{fancyhdr}
\usepackage{a4wide} % Smaller margins = more text per page.
\usepackage[hidelinks]{hyperref}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{pdflscape}
\usepackage{multirow}
\usepackage{subcaption}
\usepackage{bbm}
\usepackage{color}
\DeclareMathOperator*{\argmax}{arg\,max}
\newcommand{\ml}[1]{\textcolor{red}{ML : #1}}
\pagestyle{fancy}
\begin{document}
\author{F\'elix Gontier $^1$, Catherine Lavandier $^2$, Pierre Aumond $^{2, 3}$\\Mathieu Lagrange $^1$, Jean-Francois Petiot $^1$}
\date{
$^1$ LS2N, UMR 6004 CNRS, Ecole Centrale de Nantes, F-44321, France\\
$^2$ ETIS, UMR 8051, University of Paris Seine, University of Cergy-Pontoise, ENSEA, CNRS, F-95000, France\\
$^3$ IFSTTAR, CEREMA, UMRAE, F-44344, Bouguenais, France
}
\title{Estimation of the perceived time of presence of sources in urban acoustic environment using deep learning techniques}
\maketitle

\begin{abstract}
The impact of urban sound on human beings has often been studied from a negative point of view (sound pollution), but in the two last decades with the soundscape approach, the positive impact has been revealed (resourcing spaces). The literature shows that the recognition of sources plays a great role in the way humans are affected by sound environnements. There is thus a need for characterizing urban acoustic environments not only with sound pressure measurements but also with source-specific attributes such as the perceived time of presence. Combined, they allow us to predict the first affective dimension of the soundscape: its pleasantness. However, computationally predicting source-specific attributes from recordings of sensors is notably harder than computing sound pressure levels. This paper demonstrates, on a controlled dataset, that machine learning techniques based on state of the art neural architectures can predict the perceived time of presence of several sound sources with excellent performance, at a degree of precision that is sufficient
to predict high level concepts like the pleasantness. A corpus of simulated sound scenes is first proposed. This corpus is shown through a listening experiment to yield similar perceptual responses compared to recorded sound scenes. The availability of the contributions of the individual sound sources within the simulated sound scenes then allows us to train and validate the proposed estimation procedure.

@@

The assessment of soundscape quality through standardized perceptual descriptors has been extensively studied~\cite{viollon2000, axelsson2010, cain2013, jeon2018, aletta2016}. Aletta et al.~\cite{aletta2016} did a review of soundscape descriptors and conclude that the potential soundscape descriptors identified so far seem to converge towards a two-dimensional soundscape model of perceived affective quality where one dimension corresponds to the pleasantness of the soundscape, and the orthogonal one corresponds to its enventfulness~\cite{axelsson2010, aumond2017, delaitre2014}. In this paper, we will focus on the first affective dimension which is pleasantness. Pleasantness is mainly linked to the perceived loudness~\cite{blauert1997, jekosch2004} but it also depends on the activity of sources of interest~\cite{nilsson2007, perez2012}. While the exact taxonomy of sources used differs across studies~\cite{guastavino2007, gygi2007, salamon2014, brown2011}, three major source types are found to contribute to pleasantness in urban contexts: technological, human and natural sources~\cite{nilsson2007, axelsson2010} where traffic noise, human voices and birds calls can respectively be used as a proxy of these three types~\cite{lavandier2006, ricciardi2014, aumond2017}. In existing models, source activity can be quantified by the dominance~\cite{hong2017}, the emergence~\cite{guastavino2006} or the time of presence~\cite{ricciardi2014}. Though, the perceptual characterization of soundscapes requires subjective inputs, and thus cannot directly be applied to the production of perceptually motivated noise maps.

Leveraging the advent of the Internet of Things (IoT) and the availability of low-cost sensor networks~\cite{ardouin2018, mydlarz2017} could allow us to predict pleasantness or other perceptual factors from acoustic measurements at a reasonable cost. Though, bridging the gap between raw data issued from the sensors to reasonable estimates of perceptual attributes is yet an ongoing area of research. Ricciardi et al.~\cite{ricciardi2014} present a physical model of pleasantness using only variations of the measured sound level. Though, because these indicators are not representative of the composition of the sound scene, the model underperforms in comparison to predictions using perceptual assessments. Aumond et al.~\cite{aumond2017} explore new indicators based on spectral and temporal variations in recordings of sound scenes. These indicators are designed to model the activity of specific sound sources, specifically human voices and birds, but their discriminative properties are limited. In order to tackle this issue, the recent advances in machine learning can be useful to better approximate the contributions of sound sources relevant for perceptual attribute prediction such as pleasantness.

@@

More precisely, the contributions of this paper are three-fold:
\begin{enumerate}
  \item Provide a corpus of simulated scenes for which relevant acoustic properties are well controlled.
  \item Propose numerical means~\footnote{Code is open source and available at the companion website.} for predicting the perceived time of presence of sound sources from raw acoustic data using deep learning approaches.
  \item Demonstrate through a comparative study with state of the art approaches that the proposed method allows the prediction of pleasantness to a sufficient degree.
\end{enumerate}

First, a perceptual test is conducted in Section~\ref{sec:data} where simulated sound scenes are compared to recordings to validate the relevance of the proposed protocol for applications in \textit{in situ} sound environments. Next, Section~\ref{sec:methods} presents the design of a deep learning architecture that performs multi-label sound source recognition at relevant time scales. Time of presence estimations from this architecture are evaluated in Section~\ref{sec:results} in the context of pleasantness prediction, in comparison to existing models from perceptual and physical indicators.

@@

A corpus is first created to validate the use of simulated sound scenes in the development of time presence prediction models. It is based on recordings from one of the four soundwalks performed in~\cite{aumond2017}, including 19 locations in the 13th district of Paris, France. A classification of these scenes is proposed in~\cite{gloaguen2017} in terms of ambiance: \textit{park}, \textit{quiet street}, \textit{noisy street} and \textit{very noisy street}. For each of the 19 recordings, 45 seconds of audio in a single channel are extracted. This duration of 45 seconds is chosen to be perceptually relevant for the retrospective evaluation of sound environments. The 45~s segments are selected to represent the properties of their respective ambiances in terms of source composition, without single events overwhelming their overall perception. The extracts are then subjectively annotated in terms of the following properties:

\begin{itemize}
\item Background sources that are present throughout the whole scene, and their respective sound level considered constant.
\item Sound events characterized for each occurrence by their source type, onset-offset and event-to-background ratio (EBR) in dB.
\end{itemize}

The 45~s sound scenes are then replicated using \textit{simScene}~\cite{rossignol2015}. \textit{simScene} is a software library in Matlab that allows the simulation of sound scenes as the additive composition of sound sources, using a database of recordings associated to these sources. Here the database is constructed from excerpts of the LibriSpeech~\cite{panayotov2015} corpus for voices and Freesound\footnote{\url{https://freesound.org}} contributions for remaining sources. The original recordings from 6 locations (P1, P3, P4, P8, P15 and P18) are chosen for perceptual comparison, as they explore diverse real-life situations with respect to environment categorization. All 6 recorded and 19 replicated scenes are normalized so that their playback sound level through the restitution system is the same as measured during recording, with a range from 63.9~dB SPL to 79.4~dB SPL.

\begin{figure}[h]
    \centering
     \begin{subfigure}[t]{0.33\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/tv_pres.eps}
    \end{subfigure}%
    \begin{subfigure}[t]{0.33\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/tb_pres.eps}
    \end{subfigure}
    \begin{subfigure}[t]{0.33\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/vb_pres.eps}
    \end{subfigure}
    \caption{Physical estimations~\cite{gontier2018} of the perceived time of presence of traffic, voices and birds sources for the corpus of simulated scenes with automatically generated scenarios.}\label{fig:tvb_pres}
\end{figure}

To further study the perceptual properties of simulated scenes, the corpus is extended with 75 simulated scenes corresponding to new scenarios. Previous studies show that pleasantness is mainly influenced by the activity of three sources: \textit{traffic}, \textit{human voices} and \textit{birds}. We choose to restrict the scene composition to these 3 sources. Newly generated scenarios should cover most real-life situations while remaining perceptually plausible. Thus, statistics on background and event properties are extracted for each of the three sources from annotations presented in~\cite{gloaguen2017}. These statistics include, conditionally to the ambiance (resp. \textit{park}, \textit{quiet street}, \textit{noisy street}, \textit{very noisy street} and \textit{square}):

@@

Adjustments are made on the variance of the considered properties to better cover possible scenarios. Voice events consist mostly of read English recordings, thus their EBR is reduced to improve realism. An additional \textit{square} ambiance is added with properties derived empirically from available data. Diverse new scenarios are then generated by sampling the background and event properties distributions, resulting in 40 scenes per ambiance for a total of 200 scenes. For each scene the perceptual time of presence for the traffic, voice and bird sources is estimated using the indicators proposed in~\cite{gontier2018}. Ideally, sound scenes should cover the resulting 3-dimensional space in a homogeneous way. Though, the time of presence is not a parameter of the simulation process and thus the distribution of the scene population in this space is not controlled. As an alternative, 75 scenes are selected that maximize the minimum pairwise distance in the 3-dimensional space. This is shown in Figure~\ref{fig:tvb_pres}. The playback sound level is also conditioned to the ambiance according to typical sound levels in urban context, and ranges from 46.6~dB SPL to 77.1~dB SPL.

The experiment corpus thus totals 100 scenes: 6 recorded and 19 replicated scenes from the Grafic project, and 75 simulated scenes.

@@

A perceptual experiment is conducted where participants are asked to evaluate sound scenes in terms of 9 criteria represented by 11 point scales (0-10). These scales are presented in French and translated in this report using standard terminology. The first 5 scales relate to general properties of the scene:
\begin{itemize}
\item Pleasantness: \textit{Unpleasant - Pleasant} (\textit{D\'esagr\'eable - Agr\'eable}) - P
\item Liveliness: \textit{Inert, amorphous - Lively, eventful} (\textit{Inerte, amorphe - Anim\'e, mouvement\'e}) - L
\item Overall loudness: \textit{Quiet - Noisy} (\textit{Silencieux - Bruyant}) - OL
\item Interest: \textit{Boring, uninteresting - Stimulating, interesting} (\textit{Ennuyeux, inint\'eressant - Stimulant, int\'eressant}) - I
\item Calmness: \textit{Agitated, chaotic - Calm, peaceful} (\textit{Agit\'e, chaotique - Calme, tranquille}) - C
\end{itemize}

@@

The perception of source activity can be linked to the notions of dominance, time of presence, or emergence. We choose the perceptual time of presence, that is the percentage of time for which the corresponding source is heard, as the activity descriptor for this study. The influence of traffic on pleasantness is further separated into that of background and salient events. Thus, 4 additional questions are presented to the participants and evaluated on the same 11 point scales:
\begin{itemize}
\item Sound level of passing vehicles: \textit{Very low - Very high} (\textit{Tr\`es faible - Tr\`es fort}) - $L_{T, p}$,
\item Time of presence of Traffic, Voices, Birds: \textit{Never - Continuously} (\textit{Jamais - Continuellement}), resp. $T_{T, p}$, $T_{V, p}$ and $T_{B, p}$
\end{itemize}
where $p$ denotes a perceptual evaluation.

Prior to the test, a short verbal introduction is given to the participants and the interface is introduced to ensure that the quantities are well understood. Although the corpus is comprised of 100 sound scenes, participants only evaluate 50 scenes: all listen to the 6 recorded and 19 replicated sound scenes, then to 25 of the 75 simulated with new scenarios according to a balanced incomplete block design~\cite{dagnelie2003}. The selection of simulated scenes is done so that all scenes in the sub-corpus are evaluated by the same number of participants. All participants are first presented with the most quiet then most noisy of the recorded scenes (resp. P3 and P15). A random listening order is generated for each subject to control ordering effects for the remaining of the test. Participants can listen to each scene once, and have to answer all questions and listen to the full extract before proceeding to the next.

The scenes are played at a given sound level as discussed in Section~\ref{sec:data_corp}, through the same computer and sound card configurations. Beyerdynamics DT-990 Pro headphones are used for all participants. These headphones are calibrated in a semi-anechoic chamber where a relation between the playback sound level and the sound card output voltage is obtained for pink noise. From this information a scaling factor is applied to the sound scenes to ensure that they are heard at the desired sound level by listeners.

A total of 23 students aged from 22 to 23 yo including 16 males and 7 females at Ecole Centrale de Nantes completed the test. All gave consent prior to the experiment and reported normal hearing. With 23 subjects, the incomplete block design is not perfectly balanced.

@@

The perceptual responses of the 6 recorded and corresponding replicated scenes are first compared. Table~\ref{tab:ogrep} shows the mean differences between assessments for pairs of scenes with equivalent scenarios. Wilcoxon signed-rank tests~\cite{wilcoxon1945} are implemented for each scene and perceptual scale to outline significant differences between assessment distributions, which are shown in bold. As the data is discretely distributed, zero differences between paired samples are included using Pratt's modification of the test~\cite{pratt1959}. On the general scales (P, L, OL, I and C), all mean differences are lower than 2 points on the 11-point Likert scale. Though, significant differences are outlined that can be linked to corresponding discrepancies in source-specific parameters. The highest difference (-5.22) is found for the assessment of the time of presence of traffic in the location P4. For this location, the background traffic in the recorded scene varies along time, it is louder in the first half of the scene than in the second half. Replicating this scene using simScene imposes a constant sound level for background sources. Thus, the background traffic is louder in the replicated scene than it is in the recording for about half of its duration. To a lesser extent the same issue explains the large difference (-4) in the assessed time of presence of voices in the location P3. Discrepancies on source-specific scales can also be interpreted by the choice of isolated samples, which is semi-random and based on a high-level source taxonomy. For example, no difference is made during annotation between child or adult speech, or depending on its expressiveness. Overall, no consistent difference between the perception of recorded and replicated scenes emerges for the studied points. Thus, the proposed hypothesis that recorded and replicated scenes generate the same perceptual responses cannot be rejected.

\begin{table}[h]
\centering
\caption{Mean differences of perceptual assessments between original and replicated sound scenes. Significantly different distributions as per a Wilcoxon signed-rank test are shown in bold (n=23, p<0.05)}
\label{tab:ogrep}
%\resizebox{\columnwidth}{!}{
\begin{tabular}{ c | c c c c c c c c c }
\hline
	 & P & L & OL & I & C & $L_{T, p}$ & $T_{T, p}$ & $T_{V, p}$ & $T_{B, p}$ \\ \hline
	P1 & 0.43 & \textbf{-1.65} & \textbf{-1.04} & 0.43 & 0.13 & \textbf{-0.91} & 0.39 & \textbf{-2.09} & 0.61 \\
	P3 & 0.26 & -0.43 & 0.30 & -1 & 0.30 & 0.35 & 1.04 & \textbf{-4} & 0.22 \\
	P4 & 0.91 & 0 & \textbf{-1.83} & 0.48 & \textbf{1.30} & \textbf{-0.96} & \textbf{-5.22} & \textbf{1.43} & 0.04 \\
	P8 & 0.26 & \textbf{-1.65} & -0.87 & -0.96 & 0.65 & \textbf{-2.04} & \textbf{-0.91} & 0.09 & \textbf{-1.43} \\
	P15 & \textbf{-1.35} & 0.52 & 0.52 & \textbf{-1.17} & 0.09 & 0.61 & 0.13 & \textbf{1.96} & \textbf{-2.74} \\
	P18 & \textbf{1.13} & -0.30 & \textbf{-1.17} & -0.43 & \textbf{1.39} & -1.04 & \textbf{-1.83} & \textbf{0.83} & \textbf{1.30} \\ \hline
\end{tabular}
%}
\end{table}

@@

Discrepancies between projections of original and replicated scenes are highlighted in Figure~\ref{fig:pspace_rec} using arrows. The standard deviation of assessments is represented by ellipses for one location (P1). The projections of assessment distributions consistently overlap for all pairs of recorded and replicated scenes, available on the companion website~\footnote{URL}. This illustrates the results in Table~\ref{tab:ogrep} where perceptual assessments were not found to differ significantly overall.

\begin{figure}[h]
    \centering
    \includegraphics[width=0.9\textwidth]{figures/pca_p1.eps}
    \caption{Biplot of the principal components analysis of average assessments for the 5 general questions on the 6 recorded and 19 replicated scenes (n=25). Red dots are projections of the assessments of each scene averaged over subjects. Arrows show discrepancies between corresponding recorded and replicated scenes. For the P1 recording location the standard deviation of the projection of individual responses is shown by ellipses.}\label{fig:pspace_rec}
\end{figure}
\begin{figure}[h]
    \centering
    \includegraphics[width=0.9\textwidth]{figures/pca_sim.eps}
    \caption{Biplot of the principal components analysis of average assessments for the 5 general questions on the 75 simulated scenes (n=75). Assessments of simulated scenes are projected as red dots, and recorded and replicated scenes as blue dots.}\label{fig:pspace_sim}
\end{figure}

>>>>

\documentclass[twocolumn]{article}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Comment / uncomment for one or two column(s) fomat
%\documentclass{article}

\usepackage[modulo,switch]{lineno}
\modulolinenumbers[1]
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Comment / uncomment for showing line numbers
\linenumbers

\usepackage[latin9]{inputenc}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}

\makeatletter\@ifundefined{date}{}{\date{}}
\makeatother

\markright{\hfill Gontier {\em et al.}, p.\ }
\pagestyle{myheadings}

\paperheight297mm \paperwidth210mm
\textwidth170mm  \textheight245mm  \oddsidemargin 20mm
\evensidemargin\oddsidemargin \hoffset-22.4mm \voffset-28.4mm
\topmargin0pt \headheight20mm \headsep4mm \topskip0mm
\footskip17.5mm \columnsep7mm \arraycolsep2pt \parindent10pt
\renewcommand{\abstractname}{Summary}

\usepackage[english]{babel}
\usepackage[T1]{fontenc}

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{fancyhdr}
\usepackage[hidelinks]{hyperref}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{pdflscape}
\usepackage{multirow}
\usepackage{subcaption}
\usepackage{bbm}
\usepackage{color}
\DeclareMathOperator*{\argmax}{arg\,max}
\newcommand{\ml}[1]{\textcolor{red}{ML : #1}}
\begin{document}
\author{F\'elix Gontier $^1$, Catherine Lavandier $^2$, Pierre Aumond $^{2, 3}$\\Mathieu Lagrange $^1$, Jean-Francois Petiot $^1$}
\date{
$^1$ LS2N, UMR 6004 CNRS, Ecole Centrale de Nantes, F-44321, France\\
$^2$ ETIS, UMR 8051, University of Paris Seine, University of Cergy-Pontoise, ENSEA, CNRS, F-95000, France\\
$^3$ IFSTTAR, CEREMA, UMRAE, F-44344, Bouguenais, France
}
\title{Estimation of the perceived time of presence of sources in urban acoustic environment using deep learning techniques}
\maketitle

\begin{abstract}
The impact of urban sound on human beings has often been studied from a negative point of view (sound pollution), but in the two last decades with the soundscape approach, the positive impact has been revealed (resourcing spaces). The literature shows that the recognition of sources plays a great role in the way humans are affected by sound environments. There is thus a need for characterizing urban acoustic environments not only with sound pressure measurements but also with source-specific attributes such as their perceived time of presence. Combined, they allow us to predict important affective dimensions of the soundscape such as its pleasantness. However, computationally predicting source-specific attributes from recordings of sensors is notably harder than computing sound pressure levels.

This paper demonstrates, on a controlled dataset, that machine learning techniques based on state of the art neural architectures can predict the perceived time of presence of several sound sources with excellent performance, at a degree of precision that is sufficient
to predict perceived pleasantness. To do so, a corpus of simulated sound scenes is first designed. Perceptual attributes corresponding to those stimuli are gathered through a listening experiment. The availability of the contributions of the individual sound sources within the simulated sound scenes then allows us to train and validate the proposed estimation procedure.

% to yield similar perceptual responses compared to recorded sound scenes.
PACS numbers: 43.66.Lj (Perceptual effects of sound), 43.60.-c (Acoustic signal processing)

@@

The assessment of soundscape quality through standardized perceptual descriptors has been extensively studied~\cite{viollon2000, axelsson2010, cain2013, jeon2018, aletta2016}. Aletta et al.~\cite{aletta2016} did a review of soundscape descriptors and conclude that the potential soundscape descriptors identified so far seem to converge towards a two-dimensional soundscape model of perceived affective quality where one dimension corresponds to the pleasantness of the soundscape, and the orthogonal one corresponds to its enventfulness~\cite{axelsson2010, aumond2017, delaitre2014}. In this paper, we will focus on the first affective dimension which is pleasantness. Pleasantness is mainly linked to the perceived loudness~\cite{blauert1997, jekosch2004} but it also depends on the activity of sources of interest~\cite{nilsson2007, perez2012}. While the exact taxonomy of sources used differs across studies~\cite{guastavino2007, gygi2007, salamon2014, brown2011}, three major source types are found to contribute to pleasantness in urban contexts: technological, human and natural sources~\cite{nilsson2007, axelsson2010} where traffic noise, human voices and birds calls can respectively be used as a proxy of these three types~\cite{lavandier2006, ricciardi2014, aumond2017}. In existing models, source activity can be quantified by the dominance~\cite{hong2017}, the emergence~\cite{guastavino2006} or the time of presence~\cite{ricciardi2014}. However, the perceptual characterization of soundscapes requires subjective inputs, and thus cannot directly be applied to the production of perceptually motivated noise maps.

Leveraging the advent of the Internet of Things (IoT) and the availability of low-cost sensor networks~\cite{ardouin2018, mydlarz2017} could allow us to predict pleasantness or other perceptual factors from acoustic measurements at a reasonable cost. Though, bridging the gap between raw data issued from the sensors to reasonable estimates of perceptual attributes is yet an ongoing area of research. Ricciardi et al.~\cite{ricciardi2014} present a physical model of pleasantness using the variations of the measured sound level only. Because these indicators are not representative of the composition of the sound scene, the model underperforms when compared to the predictions computed using perceptual assessments. Aumond et al.~\cite{aumond2017} explore new indicators based on spectral and temporal variations in recordings of sound scenes. These indicators are designed to model the activity of specific sound sources, specifically human voices and birds, but their discriminative properties are limited. In order to tackle this issue, the recent advances in machine learning can be useful to better approximate the contributions of sound sources relevant for perceptual attribute prediction such as pleasantness.

@@

More precisely, the contributions of this paper are three-fold:
\begin{enumerate}
  \item Provide a corpus of simulated scenes for which relevant acoustic properties are well controlled\footnote{Corpus available at \url{} \ml{mettre sur archive}}.
  \item Propose numerical means~\footnote{Open source code is available at \url{}. \ml{mettre url}} for predicting the perceived time of presence of sound sources from raw acoustic data using deep learning approaches.
  \item Demonstrate through a comparative study with state of the art approaches that the proposed method allows the prediction of pleasantness to a sufficient degree.
\end{enumerate}

First, a perceptual test is conducted in Section~\ref{sec:data} where simulated sound scenes are compared to recordings in order to validate the relevance of the proposed protocol for \textit{in situ} sound environments. Section~\ref{sec:methods} presents the design of a deep learning architecture that performs multi-label sound source recognition at relevant time scales. Time of presence estimations from this architecture are evaluated in Section~\ref{sec:results} in the context of pleasantness prediction, in comparison to existing models from perceptual and physical indicators.

@@

A corpus is first created to validate the use of simulated sound scenes in the development of time presence prediction models. It is based on recordings from one of the four soundwalks performed in~\cite{aumond2017}, including 19 locations in the 13th district of Paris, France. A classification of these scenes is proposed in~\cite{gloaguen2017} in terms of ambiance: \textit{park}, \textit{quiet street}, \textit{noisy street} and \textit{very noisy street}. For each of the 19 recordings, 45 seconds of audio in a single channel are extracted. This duration is chosen to be perceptually relevant for the perceptual evaluation of sound environments. The 45~s segments are selected to represent the properties of their respective ambiances in terms of source composition, without single events overwhelming their overall perception. The extracts are then subjectively annotated in terms of the following properties:

\begin{itemize}
\item Background sources that are present throughout the whole scene, and their respective sound level considered constant,
\item Sound events characterized for each occurrence by their source type, onset-offset and event-to-background ratio (EBR) in dB.
\end{itemize}

The 45~s sound scenes are then replicated using the  \textit{simScene} software, an open source software library in Matlab that allows the simulation of sound scenes as the additive composition of sound sources, using a database of recordings associated to these sources\footnote{\url{https://bitbucket.org/mlagrange/simscene}}. Here the database is constructed from excerpts of the LibriSpeech~\cite{panayotov2015} corpus for voices and Freesound\footnote{\url{https://freesound.org}} contributions for remaining sources. The original recordings from 6 locations (P1, P3, P4, P8, P15 and P18) are chosen for perceptual comparison, as they explore diverse real-life situations with respect to environment categorization. The 6 recorded and 19 replicated scenes are normalized so that their playback sound level through the restitution system is the same as measured during recording, with a range from 63.9~dB SPL to 79.4~dB SPL.

\begin{figure*}[th]
    \centering
     \begin{subfigure}[t]{0.33\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/tv_pres.eps}
    \end{subfigure}%
    \begin{subfigure}[t]{0.33\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/tb_pres.eps}
    \end{subfigure}
    \begin{subfigure}[t]{0.33\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/vb_pres.eps}
    \end{subfigure}
    \caption{Physical estimations~\cite{gontier2018} of the perceived time of presence of traffic, voices and birds sources for the corpus of simulated scenes with automatically generated scenarios.}\label{fig:tvb_pres}
\end{figure*}

To further study the perceptual properties of simulated scenes, the corpus is extended with 75 simulated scenes corresponding to new scenarios. Previous studies show that pleasantness is mainly influenced by the activity of three sources: \textit{traffic}, \textit{human voices} and \textit{birds}. We choose to restrict the scene composition to these 3 sources. Newly generated scenarios should cover most real-life situations while remaining perceptually plausible. To do so, statistics on background and event properties are extracted for each of the three sources from annotations presented in~\cite{gloaguen2017}. These statistics include, conditionally to the ambiance (resp. \textit{park}, \textit{quiet street}, \textit{noisy street}, \textit{very noisy street} and \textit{square}):

@@

Adjustments are made on the variance of the considered properties to better cover possible scenarios. Since voice events consist mostly of read English recordings, thus their EBR is reduced to improve realism. An additional \textit{square} ambiance is added with properties derived empirically from available data. Diverse new scenarios are then generated by sampling the background and event properties distributions, resulting in 40 scenes per ambiance for a total of 200 scenes. For each scene the perceptual time of presence for the traffic, voice and bird sources is estimated using the indicators proposed in~\cite{gontier2018}. Ideally, sound scenes should cover the resulting 3-dimensional space in a homogeneous way. To do so, the 75 scenes that maximize the minimum pairwise distance in the 3-dimensional space are selected, as shown in Figure~\ref{fig:tvb_pres}. The playback sound level is also conditioned to the ambiance according to typical sound levels in urban context, and ranges from 46.6~dB SPL to 77.1~dB SPL.

% Though, the time of presence is not a parameter of the simulation process and thus the distribution of the scene population in this space is not controlled. As an alternative,

The experiment corpus thus totals 100 scenes: 6 recorded, 19 replicated scenes from the Grafic project, and 75 simulated scenes.

@@

In order to gather estimates of the perceputal attributes and to validate the usefulness of the proposed corpus, a perceptual experiment is conducted. Participants are asked to evaluate sound scenes in terms of 9 criteria represented by 11 point scales (0-10). These scales are presented in French and translated in this report using standard terminology. The first 5 scales relate to general properties of the scene:
\begin{itemize}
\item Pleasantness: \textit{Unpleasant - Pleasant} (\textit{D\'esagr\'eable - Agr\'eable}) - P,
\item Liveliness: \textit{Inert, amorphous - Lively, eventful} (\textit{Inerte, amorphe - Anim\'e, mouvement\'e}) - L,
\item Overall loudness: \textit{Quiet - Noisy} (\textit{Silencieux - Bruyant}) - OL,
\item Interest: \textit{Boring, uninteresting - Stimulating, interesting} (\textit{Ennuyeux, inint\'eressant - Stimulant, int\'eressant}) - I,
\item Calmness: \textit{Agitated, chaotic - Calm, peaceful} (\textit{Agit\'e, chaotique - Calme, tranquille}) - C.
\end{itemize}

@@

The perception of source activity can be linked to the notions of dominance, time of presence, or emergence. The perceptual time of presence, that is the percentage of time for which the corresponding source is heard, is selected as the activity descriptor for this study. The influence of traffic on pleasantness is further separated into that of background and salient events. Thus, 4 additional questions are presented to the participants and evaluated on the same 11 point scales:
\begin{itemize}
\item Sound level of passing vehicles: \textit{Very low - Very high} (\textit{Tr\`es faible - Tr\`es fort}) - $L_{T, p}$,
\item Time of presence of Traffic, Voices, Birds: \textit{Never - Continuously} (\textit{Jamais - Continuellement}), resp. $T_{T, p}$, $T_{V, p}$ and $T_{B, p}$
\end{itemize}
where $p$ denotes a perceptual evaluation.

Prior to the test, a short verbal introduction is given to the participants and the interface is introduced to ensure that the quantities are well understood. Although the corpus is comprised of 100 sound scenes, participants only evaluate 50 scenes: all listen to the 6 recorded and 19 replicated sound scenes, then to 25 of the 75 simulated with new scenarios according to a balanced incomplete block design~\cite{dagnelie2003}. The selection of simulated scenes is done so that all scenes in the sub-corpus are evaluated by the same number of participants. All participants are first presented with the most quiet then loudest of the recorded scenes (resp. P3 and P15). A random listening order is generated for each subject to control ordering effects for the remaining of the test. Participants can listen to each scene once, and have to listen to the full extract and to answer all questions before being allowed to proceed to the next scene.

The scenes are played at a given sound level as discussed in Section~\ref{sec:data_corp}, through the same computer and sound card configurations. Beyerdynamics DT-990 Pro headphones are used for all participants. These headphones are calibrated in a semi-anechoic chamber where a relation between the playback sound level and the sound card output voltage is obtained for pink noise. From this information a scaling factor is applied to the sound scenes to ensure that they are heard at the desired sound level by every listeners.

A total of 23 students aged from 22 to 23 years including 16 males and 7 females at Ecole Centrale de Nantes completed the test. All gave consent prior to the experiment and reported normal hearing. With 23 subjects, the incomplete block design is not perfectly balanced.

@@

The perceptual responses of the 6 recorded and corresponding replicated scenes are first studied. Table~\ref{tab:ogrep} shows the mean differences between assessments for pairs of scenes with equivalent scenarios. Wilcoxon signed-rank tests~\cite{wilcoxon1945} are implemented for each scene and perceptual scale to outline significant differences between assessment distributions, which are shown in bold. As the data is discretely distributed, zero differences between paired samples are included using Pratt's modification of the test~\cite{pratt1959}. On the general scales (P, L, OL, I and C), all mean differences are lower than 2 points on the 11-point Likert scale. Though, significant differences are outlined that can be linked to corresponding discrepancies in source-specific parameters. The highest difference (-5.22) is found for the assessment of the time of presence of traffic in the location P4. For this location, the background traffic in the recorded scene varies along time, it is louder in the first half of the scene than in the second half. Replicating this scene using simScene imposes a constant sound level for background sources. Thus, the background traffic is louder in the replicated scene than it is in the recording for about half of its duration. To a lesser extent the same issue explains the large difference (-4) in the assessed time of presence of voices in the location P3. Discrepancies on source-specific scales can also be interpreted by the choice of isolated samples, which is semi-random and based on a high-level source taxonomy. For example, no difference is made during annotation between child or adult speech, or depending on its expressiveness. Overall, no consistent difference between the perception of recorded and replicated scenes emerges for the studied points. Thus, the proposed hypothesis that recorded and replicated scenes generate equivalent perceptual responses cannot be rejected.

\begin{table*}[t]
\centering
\caption{Mean differences of perceptual assessments between original and replicated sound scenes. Significantly different distributions as per a Wilcoxon signed-rank test are shown in bold (n=23, p<0.05)}
\label{tab:ogrep}
%\resizebox{\columnwidth}{!}{
\begin{tabular}{ c | c c c c c c c c c }
\hline
	 & P & L & OL & I & C & $L_{T, p}$ & $T_{T, p}$ & $T_{V, p}$ & $T_{B, p}$ \\ \hline
	P1 & 0.43 & \textbf{-1.65} & \textbf{-1.04} & 0.43 & 0.13 & \textbf{-0.91} & 0.39 & \textbf{-2.09} & 0.61 \\
	P3 & 0.26 & -0.43 & 0.30 & -1 & 0.30 & 0.35 & 1.04 & \textbf{-4} & 0.22 \\
	P4 & 0.91 & 0 & \textbf{-1.83} & 0.48 & \textbf{1.30} & \textbf{-0.96} & \textbf{-5.22} & \textbf{1.43} & 0.04 \\
	P8 & 0.26 & \textbf{-1.65} & -0.87 & -0.96 & 0.65 & \textbf{-2.04} & \textbf{-0.91} & 0.09 & \textbf{-1.43} \\
	P15 & \textbf{-1.35} & 0.52 & 0.52 & \textbf{-1.17} & 0.09 & 0.61 & 0.13 & \textbf{1.96} & \textbf{-2.74} \\
	P18 & \textbf{1.13} & -0.30 & \textbf{-1.17} & -0.43 & \textbf{1.39} & -1.04 & \textbf{-1.83} & \textbf{0.83} & \textbf{1.30} \\ \hline
\end{tabular}
%}
\end{table*}

@@

Discrepancies between projections of original and replicated scenes are highlighted in Figure~\ref{fig:pspace_rec} using arrows. The standard deviation of assessments is represented using ellipses for one location (P1). The projections of assessment distributions consistently overlap for all pairs of recorded and replicated scenes\footnote{Data available here \url{} \ml{mettre url}}. This illustrates the results in Table~\ref{tab:ogrep} where perceptual assessments were not found to differ significantly overall.

\ml{On vise 12 pages. Tu dois raccourcir les legendes des figures.}

\begin{figure}[th]
    \centering
    \includegraphics[width=\columnwidth]{figures/pca_p1.eps}
    \caption{Biplot of the principal components analysis of average assessments for the 5 general questions on the 6 recorded and 19 replicated scenes (n=25). Red dots are projections of the assessments of each scene averaged over subjects. Arrows show discrepancies between corresponding recorded and replicated scenes. For the P1 recording location the standard deviation of the projection of individual responses is shown using ellipses.\ml{Pas de couleur, Remplacer par des croix et des points !! enlever le blanc autour de la figure, soit a la main, soit en controlant la bounding box}}\label{fig:pspace_rec}
\end{figure}
\begin{figure}[h]
    \centering
    \includegraphics[width=\columnwidth]{figures/pca_sim.eps}
    \caption{Biplot of the principal components analysis of average assessments for the 5 general questions on the 75 simulated scenes (n=75). Assessments of simulated scenes are projected as red dots, and recorded and replicated scenes as blue dots.}\label{fig:pspace_sim}
\end{figure}
